---
title: "exploring decontaminated metabarcoding reads"
author: "Kimberly Ledger"
date: "2024-07-25"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

libraries
```{r}
library(tidyverse)
library(ggplot2)
rename <- dplyr::rename
```


the data 
```{r}
taxon_table <- read.csv("/home/kimberly.ledger/BeringSea_Arctic_eDNA/outputs/20240806/decontaminated_taxon_table.csv") %>%
  select(!X)

metadata <- read.csv("/home/kimberly.ledger/BeringSea_Arctic_eDNA/data/NBS_SBS_DBO_metadata.csv") 

metadata_mini <- metadata %>%
  select(extraction_ID, alternative_ID, sample_type, project, collection_year, location1, location2, depth, longitude, latitude) %>%
  unite("project_year", project, collection_year, sep = "_", remove = F)

# for DBO2021 samples, replace extraction_ID with alternative_ID
metadata_mini <- metadata_mini %>%
  mutate(extraction_ID = ifelse(project_year == "DBO_2021", alternative_ID, extraction_ID)) %>%
  select(!alternative_ID)
```

export station numbers for NBS/SBS
```{r}
#projects <- c("NBS", "SBS")

# temp <- metadata_mini %>%
#   filter(project %in% projects) %>%
#   select(project, collection_year, location1, depth, longitude, latitude) %>%
#   filter(!is.na(location1)) %>%
#   filter(!is.na(depth)) %>%
#   unique()

#write.csv(temp, "/home/kimberly.ledger/BeringSea_Arctic_eDNA/outputs/stations.csv")
```



for now, i'm going to just work with field samples (i.e. ignore any reads in remaining field negatives, etc)
```{r}
taxon_table_meta <- metadata_mini %>%
  filter(sample_type == "sample") %>%
  left_join(taxon_table)
```

export to use with 
```{r}
write.csv(taxon_table_meta, "../outputs/20240806/taxon_table_w_meta.csv")
```


make some plots for DBO 2021

spp in DBO 2021
```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2021") %>%
  filter(tot_reads > 0) %>%
  group_by(taxon) %>%
  summarize(n_reads = sum(tot_reads)) %>%
  arrange(desc(n_reads))
```
```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2021") %>%
  filter(location2 == "DBO_1") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 1 - 2021") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2021") %>%
  filter(location2 == "DBO_2") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 2 - 2021") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2021") %>%
  filter(location2 == "DBO_3") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 3 - 2021") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2021") %>%
  filter(location2 == "DBO_4") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 4 - 2021") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2021") %>%
  filter(location2 %in% c("CEO", "C2", "M8")) %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Moorings - 2021") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

make some plots for DBO 2023

spp in DBO 2023 
```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(tot_reads > 0) %>%
  group_by(taxon) %>%
  summarize(n_reads = sum(tot_reads)) %>%
  arrange(desc(n_reads))
```


```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_1") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 1 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r, fig.width=7.5, fig.height=5}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  #filter(location2 == "DBO_1") %>%
  filter(location1 == "DBO_1.4") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "depth",
    title = "DBO Station 1 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_1") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  group_by(taxon) %>%
  summarize(mean_prop = mean(prop)) %>%
  arrange(desc(mean_prop))
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_2") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 2 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r, fig.width=8.5, fig.height=5}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_2") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "depth",
    title = "DBO Station 2 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_2") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  group_by(taxon) %>%
  summarize(mean_prop = mean(prop)) %>%
  arrange(desc(mean_prop))
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_3") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 3 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r, fig.width=8.5, fig.height=5}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_3") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "depth",
    title = "DBO Station 3 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```


```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_3") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  group_by(taxon) %>%
  summarize(mean_prop = mean(prop)) %>%
  arrange(desc(mean_prop))
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_4") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 4 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```


```{r, fig.width=8.5, fig.height=5}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_4") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "depth",
    title = "DBO Station 4 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_4") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  group_by(taxon) %>%
  summarize(mean_prop = mean(prop)) %>%
  arrange(desc(mean_prop))
```


```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_5") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 5 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r, fig.width=8.5, fig.height=5}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_5") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "depth",
    title = "DBO Station 5 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_5") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  group_by(taxon) %>%
  summarize(mean_prop = mean(prop)) %>%
  arrange(desc(mean_prop))
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 %in% c("BF2", "CK9", "M8")) %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Moorings - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```



now let's plot relative read abundance 
```{r}
taxon_table_meta %>%
  unite(col = "station_depth", location1, depth, sep = "_", remove = F) %>%
  filter(project_year == "NBS_2021") %>%
  filter(location1 %in% c("2", "8", "11", "17")) %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  ggplot(aes(x=extraction_ID, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~station_depth, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "sample ID",
    title = "NBS 2021") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```


run an NMDS comparing DBO stations in a given year and then also across years(?) 
- since the % reads have not been corrected for any PCR biases, i will air on the side of caution and just input presence/absence data 


summarize the number of detections per taxon
```{r}
taxon_table_meta_w_detections <- taxon_table_meta %>%
  filter(project == "DBO") %>%
  mutate(present = ifelse(tot_reads > 0, 1, 0)) %>%
  filter(present == 1) %>%
  group_by(collection_year, taxon) %>%
  mutate(n_detections_in_year = sum(present),
         n_reads_in_year = sum(tot_reads)) %>%
  arrange(collection_year, n_detections_in_year)
```

```{r}
dbo_pres <- taxon_table_meta_w_detections %>%
  filter(project == "DBO") %>%
  filter(n_detections_in_year > 2) %>%
  filter(n_reads_in_year > 99) %>%
  #mutate(present = ifelse(tot_reads > 0, 1, 0)) %>%
  #filter(present == 1) %>%
  #select(collection_year, location1, location2, depth, taxon, present) %>%
  #pivot_wider(names_from = taxon, values_from = present, values_fill = 0)
  select(collection_year, location1, location2, depth, longitude, latitude, taxon, tot_reads) %>%
  pivot_wider(names_from = taxon, values_from = tot_reads, values_fill = 0)
```


```{r}
library(vegan)
```

```{r}
# dbo_pres_21 <- dbo_pres %>%
#   filter(collection_year == 2021)
# 
# matrix_21 <- dbo_pres_21[,-c(1:4)]
# meta_21 <- dbo_pres_21[,c(1:4)]
# 
# dbo_pres_23 <- dbo_pres %>%
#   filter(collection_year == 2023)
# 
# matrix_23 <- dbo_pres_23[,-c(1:4)]
# meta_23 <- dbo_pres_23[,c(1:4)]

matrix <- dbo_pres[,-c(1:6)]
meta <- dbo_pres[,c(1:6)]
```


```{r}
#NMS_data<-metaMDS(matrix_21, distance = "jaccard", k=2, try=100, autotransform = TRUE, maxit=1000)
#NMS_data<-metaMDS(matrix_21, distance = "bray", k=2, try=100, autotransform = TRUE, maxit=1000)

#NMS_data<-metaMDS(matrix_23, distance = "bray", k=2, try=100, autotransform = TRUE, maxit=1000)

#NMS_data<-metaMDS(matrix, distance = "jaccard", k=2, try=100, autotransform = TRUE, maxit=1000)
NMS_data<-metaMDS(matrix, distance = "bray", k=2, try=100, autotransform = TRUE, maxit=1000)
```

```{r}
#create vectors with the NMS attributes
NMS_coordinates<-scores(NMS_data,display="sites")
NMS_axes<-as.data.frame(NMS_coordinates)
NMS_scores<-scores(NMS_data,display="species")
```

link back sample metadata
```{r}
#for_ploting<-as.data.frame(cbind(NMS_coordinates,meta_21))
#for_ploting<-as.data.frame(cbind(NMS_coordinates,meta_23))
for_ploting<-as.data.frame(cbind(NMS_coordinates,meta))
```

```{r}
#plot the stress
stressplot(NMS_data)

NMS_data$stress
```

```{r}
nmds.plot <- ggplot(for_ploting, aes(x=NMDS1, y=NMDS2))+ #sets up the plot
  #geom_point(aes(NMDS1, NMDS2, colour = factor(for_ploting$location2), shape = factor(for_ploting$collection_year)), size = 2)+ #adds site points to plot, shape determined by depth, colour determined by location
  geom_point(aes(NMDS1, NMDS2, colour = for_ploting$latitude, shape = factor(for_ploting$collection_year)), size = 2)+ #adds site points to plot, shape determined by depth, colour determined by location
  coord_fixed()+
  theme_classic()+ 
  theme(panel.background = element_rect(fill = NA, colour = "black", size = 1, linetype = "solid"))+
  #labs(colour = "Station", shape = "Year") + # add legend labels for Station
  labs(colour = "Latitude", shape = "Year", title = "bray") + # add legend labels for Station
  theme(legend.position = "right", 
        legend.text = element_text(size = 12), 
        legend.title = element_text(size = 12), 
        axis.text = element_text(size = 10)) #+ # add legend at right of plot
  #xlim(-0.01,0.015) + 
  #ylim(-0.01,0.01)

nmds.plot
```

```{r}
#spp.fit <- envfit(NMS_data, matrix_21, permutations = 999) # this fits species vectors
#spp.fit <- envfit(NMS_data, matrix_23, permutations = 999) # this fits species vectors
spp.fit <- envfit(NMS_data, matrix, permutations = 999) # this fits species vectors

spp.scrs <- as.data.frame(scores(spp.fit, display = "vectors")) #save species intrinsic values into dataframe
spp.scrs <- cbind(spp.scrs, Species = rownames(spp.scrs)) #add species names to dataframe
spp.scrs <- cbind(spp.scrs, pval = spp.fit$vectors$pvals) #add pvalues to dataframe so you can select species which are significant
#spp.scrs<- cbind(spp.scrs, abrev = abbreviate(spp.scrs$Species, minlength = 6)) #abbreviate species names
sig.spp.scrs <- subset(spp.scrs, pval<=0.05) #subset data to show species significant at 0.05

head(spp.scrs)

library(ggrepel)
# nmds.plot +
#   geom_segment(data = sig.spp.scrs, aes(x = 0, xend=NMDS1, y=0, yend=NMDS2), arrow = arrow(length = unit(0.25, "cm")), colour = "grey10", lwd=0.3) + #add vector arrows of significant species
#   ggrepel::geom_text_repel(data = sig.spp.scrs, aes(x=NMDS1, y=NMDS2, label = Species), cex = 3, direction = "both", segment.size = 0.25)


#plot just significant species 
nmds.plot.spp <- ggplot() + 
  geom_point(data = sig.spp.scrs, aes(NMDS1, NMDS2, colour = Species)) + 
  ggrepel::geom_text_repel(data = sig.spp.scrs, aes(x=NMDS1, y=NMDS2, label = Species), cex = 3, direction = "both", segment.size = 0.25) + 
  theme_classic()+ 
  theme(legend.position = "none") +
  labs(title = "significant species")

nmds.plot.spp   
```



```{r}
taxon_table_meta %>%
  filter(collection_year == 2021) %>%
  filter(location1 == "DBO_4.4") %>%
  filter(depth == 30) %>%
  filter(tot_reads > 0)
```

```{r}
taxon_table_meta %>% 
  filter(project == "DBO") %>%
  filter(taxon == "Allosmerus elongatus") %>%
  filter(tot_reads > 0)
```

Icelus only in one sample
remove all species with only one (or few detections?)
remove taxa with very very few reads (ie. Allosmerus elongatus)

*those two species still causing problems... 

is there a clear latitude cutoff/change point in species comps? 

```{r}
meta %>%
  filter(latitude > 70) %>%
  select(location2) %>%
  unique()

meta %>%
  filter(latitude < 70) %>%
  select(location2) %>%
  unique()
```

```{r}
taxon_table_meta %>%
  filter(project == "DBO") %>%
  filter(latitude > 70) %>%
  filter(tot_reads > 0) %>%
  ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  #ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(collection_year~location2, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO above 70N") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r, fig.width=8.5, fig.height=5}
taxon_table_meta %>%
  filter(project == "DBO") %>%
  filter(latitude > 70) %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  ggplot(aes(x=extraction_ID, y=prop, fill=taxon)) +
  #ggplot(aes(x=depth, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(collection_year~location2, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "depth",
    title = "DBO above 70N") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project == "DBO") %>%
  filter(latitude > 70) %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  group_by(taxon) %>%
  summarize(mean_prop = mean(prop)) %>%
  arrange(desc(mean_prop))

taxon_table_meta %>%
  filter(project == "DBO") %>%
  filter(latitude > 70) %>%
  filter(tot_reads > 0) %>%
  group_by(taxon) %>%
  summarize(total_reads = mean(tot_reads)) %>%
  arrange(desc(total_reads))
```




```{r, fig.width=8.5, fig.height=5}
taxon_table_meta %>%
  filter(project == "DBO") %>%
  filter(latitude < 70) %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  ggplot(aes(x=extraction_ID, y=prop, fill=taxon)) +
  #ggplot(aes(x=depth, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(collection_year~location2, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "depth",
    title = "DBO belove 70N") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project == "DBO") %>%
  filter(latitude < 70) %>%
  filter(tot_reads > 0) %>%
  group_by(taxon) %>%
  summarize(total_reads = mean(tot_reads)) %>%
  arrange(desc(total_reads))
```

