---
title: "field sample stats - accumulation curves"
author: "Kimberly Ledger"
date: "2024-09-27"
output: html_document
---


libraries
```{r}
library(tidyverse)
rename <- dplyr::rename
```

input decontaminated ASV table
```{r}
asv_table <- read.csv("/home/kimberly.ledger/BeringSea_Arctic_eDNA/outputs/20240806/decontaminated_asv_table.csv") %>%
  select(!X) %>%
  filter(project == "DBO") 

asv_table_2 <- asv_table %>%
  separate(sample_ID_date, into = c("extraction_ID", "replicate", "seq_date2"), sep = "-")
```


## explore the field samples 
- use species accumulation curves with respect to the number of corrected reads to explore if that the sequencing depth was sufficient to detect all of the species contained in a sample

since i only have one PCR replicate for 2021 samples, i'm going to switch from filtering by the pcr replicate level, and going to consider reads per bottle of water (extraction_ID / biological replicate)- also going to work at the taxon level, instead of with ASVs 

```{r}
taxon_table_1 <- asv_table_2 %>%
  group_by(extraction_ID, project_year, project, collection_year, sample_type, taxon, taxonomic_level, species, genus, family, order, class) %>%
  summarize(tot_reads = sum(reads))
```

let's get to the accumulation curve using extraction IDs 

read in some metadata that groups extractions IDs into sites 
```{r}
metadata <- read.csv("/home/kimberly.ledger/BeringSea_Arctic_eDNA/data/NBS_SBS_DBO_metadata.csv") %>%
  filter(project == "DBO")

metadata_mini <- metadata %>%
  select(extraction_ID, alternative_ID, sample_type, project, collection_year, location1, location2, depth, longitude, latitude) %>%
  unite("project_year", project, collection_year, sep = "_", remove = F)

# for DBO2021 samples, replace extraction_ID with alternative_ID
metadata_mini <- metadata_mini %>%
  mutate(extraction_ID = ifelse(project_year == "DBO_2021", alternative_ID, extraction_ID)) %>%
  select(!alternative_ID)
```


try out accumulation curve to evaluate sequencing depth 
```{r}
library(iNEXT)
library(ggplot2)

#iNext crashes easily because of too many samples.  will need to break this up by project and year and subsets of samples...
DBO_2021 <- taxon_table_1 %>%
filter(collection_year == 2021)


# need N lists of species read counts 

taxon_table_2 <- DBO_2021 %>%
   pivot_wider(names_from = taxon, values_from = tot_reads)

# now convert the taxon table to lists of species abundances (one list per pcr rep)

# make a variable for the first and last ASV column in the table
tax_first <- which(colnames(taxon_table_2) == "Acantholumpenus mackayi")  
tax_last <- ncol(taxon_table_2)

# okay, need to go back to long format 
taxon_long <- taxon_table_2 %>%
  pivot_longer(cols = tax_first:tax_last, names_to = "taxon", values_to = "reads") %>%    
  filter(reads > 0) %>% # filter out entries with no reads
  arrange(extraction_ID, desc(reads))  # sort reads in descending order and group by sample

#st <- taxon_long %>%
#  filter(location1 %in% c("5"))

# Nest the data to create a list of species reads for each sample
read_lists <- taxon_long %>%
  #group_by(sample_ID_date) %>%
  group_by(extraction_ID) %>%
  nest() %>%
  mutate(
    species_read_list = map(data, ~ .x %>% select(taxon, reads))
  ) %>%
  #select(sample_ID_date, species_read_list)
  select(extraction_ID, species_read_list)

#Convert to a list of lists
read_list_final <- read_lists %>%
  pull(species_read_list) %>%
  #setNames(read_lists$sample_ID_date)
  setNames(read_lists$extraction_ID)

# Convert each table to a numeric vector of abundances
read_lists_vector <- map(read_list_final, ~ {
  .x %>%
    filter(reads > 0) %>%
    arrange(desc(reads)) %>%
    pull(reads)
})

# knots is the number of intervals
# endpoint is the max read count 
# m can designate sample size 

# try to run iNEXT - the function doesn't like non-integers so lots of warning will pop up.
#out_nbs2023_st38 <- iNEXT(read_lists_vector, q = 0, datatype = "abundance")
#out_nbs2023_st2_80knots <- iNEXT(read_lists_vector, q = 0, datatype = "abundance", knots = 80, endpoint = 100000) 
#out_nbs2021_st46 <- iNEXT(read_lists_vector, q = 0, datatype = "abundance", knots = 80)

#set a series of sample sizes for (m) 
m <- c(1,50,100,250,500,1000,2000,3000,4000,5000,10000,20000,30000,40000)

out_dbo21_m <- iNEXT(read_lists_vector, q = 0, datatype = "abundance", size = m)
#out_nbs2021_st5_m <- iNEXT(read_lists_vector, q = 0, datatype = "abundance", size = m)

```

check out the output 
```{r}
out_nbs2021_st20_m$DataInfo
out_nbs2021_st20_m$iNextEst
out_nbs2021_st20_m$AsyEst
```
 
custom plot since ggiNEXT is giving me color problems 
```{r}
df <- fortify(out_nbs2021_st5_m, type=1)
df.point <- df[which(df$Method=="Observed"),]
df.line <- df[which(df$Method!="Observed"),]
df.line$Method <- factor(df.line$Method,  c("Rarefaction", "Extrapolation"), c("Rarefaction", "Extrapolation"))

my_plot <- ggplot(df, aes(x=x, y=y, colour=Assemblage)) +  
  geom_point(size=5, data=df.point) + 
  geom_line(aes(linetype=Method), lwd=1.5, data=df.line) + 
  geom_ribbon(aes(ymin=y.lwr, ymax=y.upr, fill=Assemblage, colour=NULL), alpha=0.2) + 
  labs(x="number of sequencing reads per 1-L bottle", y="Number of detected taxa", 
       title = "NBS 2021 Station 5") + 
  theme(legend.position = "bottom",  legend.title=element_blank(), text=element_text(size=18), legend.box = "vertical") +
  xlim(0, 10000) + 
  theme_minimal()
my_plot
```
 
```{r}
#ggsave("/home/kimberly.ledger/BeringSea_Arctic_eDNA/outputs/accumulation_curves/NBS_2023_st12_23_38_43_zoom.png", plot = my_plot, dpi = 150, width = 8, height = 6)
```

```{r}
#extract iNEXT result 
accumulation_data <- out_nbs2021_st5_m$iNextEst[[1]] 
```

maybe try to come back see if i can quantitatively determine the number of reads to be within x % of the max/observed species richness???? 



###############  okay trying a different way to do accumulation curves 

```{r}
library(vegan)
library(ggplot2)
```

```{r}
taxon_table_wide <- taxon_table_1[,c(1,6,13)] %>%
  mutate(tot_reads = as.integer(tot_reads)) %>%
  pivot_wider(names_from = taxon, values_from = tot_reads)

extraction_IDs <- taxon_table_wide$extraction_ID
taxon_table_wide <- taxon_table_wide[,-1]

my_rarecurve <- rarecurve(taxon_table_wide, step = 20, col = "blue", label = FALSE, 
          main = "Sequencing Effort Curves",
          xlab = "Sequencing Depth", ylab = "Number of Species Identified")
my_rarecurve
```

it looks like around 25,000 reads per sample was enough to saturate taxonomic diversity. 


this finds the number of species for a certain number of sampled sites - would likely need to break down sites by year and location to make this information useful for my project
```{r}
accum_curve <- specaccum(taxon_table_wide, "random")
summary(accum_curve)
plot(accum_curve)
```

what is the total reads per sampling station (1 PCR rep per extraction for 2021; 3 combined PCR reps per extraction in 2023)

```{r}
reads <- taxon_table_1 %>%
  group_by(extraction_ID, project_year, sample_type) %>%
  summarize(reads = sum(tot_reads)) %>%
  arrange(desc(reads))

ggplot(reads, aes(x = reorder(extraction_ID, -reads), y = reads)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(title = "Total Reads per Sample (Ordered from Largest to Smallest)",
       x = "Sample",
       y = "Total Reads") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels for readability
```

summarize in a table how many bottles meet certain read count thresholds 
```{r}
read_summary <- reads %>%
  group_by(project_year, sample_type) %>%
  summarize(atleast1 = sum(reads >= 1),
            atleast500 = sum(reads >= 500),
            atleast1k = sum(reads >= 1000),
            atleast2k = sum(reads >= 2000),
            atleast5k = sum(reads >= 5000),
            atleast10k = sum(reads >= 10000),
            atleast20k = sum(reads >= 20000))
```

### continue here.  need to pick a per sample read threshold. 

### resumed here. 

for now, i am going to aggregate data by the extraction ID and remove any extractions with less than 2000 reads total from the dataset. 

```{r}
bottles_lessthan_2000 <- taxon_table_1 %>%
  group_by(extraction_ID, project_year, sample_type) %>%
  summarize(reads = sum(tot_reads)) %>%
  filter(reads < 2000)
```

```{r}
taxon_table_3 <- taxon_table_1 %>%
  filter(!extraction_ID %in% bottles_lessthan_2000$extraction_ID)
```

```{r}
#write.csv(taxon_table_1, "/home/kimberly.ledger/BeringSea_Arctic_eDNA/outputs/20240806/decontaminated_taxon_table.csv")
#write.csv(taxon_table_3, "/home/kimberly.ledger/BeringSea_Arctic_eDNA/outputs/20240806/decontaminated_taxon_table_removedbottleslessthan2000.csv")
```



